Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-01 paper.tex
Yans updates
Clear up some author comments

% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real time.
Building on the open-source \texttt{tskit} library, the method
explicitly reconstructs genealogies with both
mutation and recombination, which
can be conveniently and efficiently analysed using standard data science
tools. Inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases.
% First, likely attachment nodes in the current ARG
% (or paths, if inferred to be a recombinant) are
First, possible paths connecting each sample to the current ARG
are inferred (allowing for recombination) using the
% chosen for each of the samples in the daily batch using the
Li and Stephens (LS) model (Section~\ref{sec:ls}). Then, since many samples
in the batch can share an attachment path,
we infer phylogenetic trees for each of the clusters separately using standard
methods (Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony-based heuristics to address issues
introduced by the inherent greediness of this strategy
(Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (with 1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of accurate joint estimation of
genetic inheritance with mutation and recombination events will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine in every generation,
in SARS-CoV-2 recombination is relatively rare, with recombination nodes accounting
for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes, e.g., restricted by time or limited to close
relatives. We expect ARGs to be particularly tree-like
early in the pandemic, when co-infection was less likely and divergence between
lineages relatively low.

@@

% Paragraph 4, what do we say about the divergence times?
Figure~\ref{fig:recomb_mrcas} shows that there is a
large range in divergence times among recombinant parents, with a broad
spread of times from about 10 to 80 weeks prior to the recombination event
and a minor additional peak of recombinants that diverged
${\sim}95$ weeks, corresponding to recombinants whose
parental common ancestor traces to early 2020. We would expect, for example
Delta-Omicron recombinants to fall into this group, hence it can be
useful to classify the breakpoints further by the VoC combinations of their
parent lineages. Considering only the Alpha, Delta, and Omicron VoC classes,
such a classification reveals that the majority of breakpoints have two Delta or
two Omicron parents, and that Omicron and Delta are the variants associated
with the most recombination (Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
This may reflect either sampling intensity, the
prevalence of cases (which increases the
chance of coinfection and recombination), or possible heterogeneity in
recombination probabilities among lineages.

@@

\section{Discussion}
% PARA 1. SARS-CoV-2 is still important, and the vast hordes of data provide an
% unprecedented opportunity to study the evolution of a virus. Even if the
% public health emergency is officially over, it doesn't mean it's not coming
% back and we should be prepared. Recombination is now a vital factor, and
% present a way in which it can be incorporated. Our results are accurate.
Although the COVID-19 pandemic is no longer considered a global emergency
by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide. This fuels
proliferation of many variants, with more than 600 Pango-designated lineages
circulating globally in the last three months (January to March, 2023;
https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
higher risk of coinfection, increasing opportunities for new phenotypically
distinct recombinants to emerge and spread. Phylogenetic approaches have
been central to pandemic responses thus far [FIXME: Examples and citations].
% 1. Form the basis of variant classifications (UShER paper),
% 2. Global context. In 2020, Nextstrain provided situation reports that were interactive, phylogenetically-driven visualizations of SARS-CoV-2 characteristics such as transmission https://nextstrain.org/ncov-sit-reps/
% 3. The WHO's recommendations on Omicron asked countries to "report initial cases/clusters associated with VOC infection to WHO". Phylogenetics being one way to identify those crucial early clusters. https://www.who.int/news/item/26-11-2021-classification-of-omicron-(b.1.1.529)-sars-cov-2-variant-of-concern
However, as recombinant lineages are now endemic and it is clear that
recombination is a major feature of ongoing SARS-CoV-2 evolution,
it is imperative that these methods are updated to incorporate the
effects of recombination, and future public health interventions are
not based on incomplete and potentially biased evolutionary models.
Here we have introduced the first method to infer an evolutionary
history that jointly estimates genealogies with both mutation and
recombination at pandemic scale, and illustrated how this single
structure accurately captures results derived by many different means.

@@

% We introduce an ARG inference method, summarised in
% Figure~\ref{fig:overview_sc2ts}.
% % Note going to rewrite this anyway, so not trying to make it flow.
% It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
% \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
% the reported collection dates of the SARS-CoV-2 samples to determine the order
% of genome copying; and (3) several parsimony-based heuristics to locally refine
% the topology. In brief, this combined approach uses the likelihood model of LS
% to attach sample genomes to existing nodes and to detect recombinants, but
% because many of the daily tip samples harbour few, if any, mutational changes,
% we use parsimony to resolve their close relationships.

% Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
% taking advantage of the reported collection dates to incrementally grow the
% ARG. The inference process occurs forward in time, attaching new genomes
% aggregated by their collection date (daily batches), from the earliest to the
% last collection date. Prior to attaching sample genomes, the ARG is initialised
% such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
% of recombination, the ARG is a single tree with no reticulation (Figure 1A);
% however, in the presence of recombination, a reticulate ARG is generated, with
% variation in tree topology across the genome (Figure 1B). Unlike standard
% phylogenetic trees in which internal nodes always represent inferred ancestors,
% internal nodes in the sc2ts ARGs represent sequences which can either be
% inferred ancestors or known samples.

% Each daily batch of genomes is first attached to the growing ARG without any
% phylogenetic resolution. In cases where the newly added sequence is inferred to
% be a recombinant, a recombination node is inserted; the ``date'' of this node
% is arbitrarily set to the average of the collection dates of the genome and the
% most recent of its parent nodes. Otherwise, the genomes are attached to the
% most closely related ancestral node (an attachment node). This creates
% polytomies that are then locally resolved among the descendants of each
% attachment node using the Unweighted Pair Group Method with Arithmetic Mean
% (UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
% This hierarchical clustering algorithm produces a dendrogram relating a set of
% sequences, which is treated as an ultrametric binary tree here. Subsequently,
% mutations are placed onto the daily tree by maximum parsimony. An issue with
% the UPGMA algorithm is that it may create internal nodes supported by no
% informative site (i.e., having no mutation immediately ancestral to them). In
% an adjustment step, such unsupported internal nodes are removed to
% ``un-resolve'' the daily tree; the remaining internal nodes, which are
% supported by at least one shared mutation, are retained. The ``dates'' of the
% non-sample internal nodes are set to the same date as the samples in the daily
% batch. The resulting daily tree is grafted onto the attachment node in the
% global ARG.

% During ARG inference, two types of non-sample nodes are inserted to improve
% parsimonious placement of mutations: mutation-coalescing nodes and
% pushed-reversion nodes. We call these operations tree adjustments. A
% mutation-collapsing node is inserted when a shared mutation between sibling
% nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
% made after grafting a new daily tree onto the global ARG. If shared mutations
% exist between two or more sibling nodes, then a parsimonious explanation is to
% create a new node that is the parent of those siblings, on which these
% mutations occurred. A pushed-reversion node is inserted when an immediate
% reversion occurs (Figure 1E), i.e., a mutation that reverts another mutation on
% its parent node. Instead of retaining both the mutation and immediate
% reversion, only the mutation is kept after a pushed-reversion node is inserted,
% thereby reducing the number of mutations on the branches. Note that these
% non-sample nodes are used as reconstructed genomes during HMM copying.

% Furthermore, during inference, the ARG is automatically encoded, stored, and
% updated as a tree sequence using tskit \citep{Kelleher2018-xc}.

@@

The LS HMM machinery, and the interpretation of inferred recombinant paths and
breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
ways to extend and improve. For example, the current parameterisation of using
a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}), and likely
results in a flat likelihood space where many recombinant paths have equal
probability of being chosen.
More sophisticated parameterisations where we
condition on known per-site mutation rates (and perhaps per-base subsitution
rates) may lead to better results, although perhaps at the cost of some
reduction in interpretability. Post-processing the match results to produce
more parsimonious breakpoints may also be a worthwhile avenue for development.
In particular, we may choose to insert breakpoints for the ARG that are chosen
from within the possible interval, rather than the current approach of taking
the rightmost value. Cases where we have more than two parents may either be
the ``stacking'' of multiple recombination events, or instances where
the HMM has chosen to switch to a third sequence rather than back to
the original parent (where this is equally parsimonious). Many putative
recombination events, however, will represent poor quality data, where
a recombinant copying path happens to be a more likely explanation
of a highly divergent sequence.
A thorough analysis of the behaviour of the LS model in the context
of a pandemic-scale ARG may lead to significant improvements in our
ability identify recombinants and to filter poor quality data.

@@

\section{Data Availability}
The source code for \texttt{sc2ts} and notebooks and code used to produce the
results described here available on GitHub:
\begin{itemize}
\item \url{https://github.com/jeromekelleher/sc2ts/}
\item\url{https://github.com/jeromekelleher/sc2ts-paper/}
\end{itemize}

@@

%% Could insert the Pango_X_graph_sample_maps.txt with something like this
%% but maybe best to leave it as a separate file?
%% \lstinputlisting[caption={XXX}]{data/Pango_X_graph_sample_maps.txt}

>>>>

% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
\usepackage[hidelinks]{hyperref}

@@

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real time.
Building on the open-source \texttt{tskit} library, the method
explicitly reconstructs genealogies with both
mutation and recombination, which
can be conveniently and efficiently analysed using standard Python data science
tools. Inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases.
% First, likely attachment nodes in the current ARG
% (or paths, if inferred to be a recombinant) are
First, possible paths connecting each sample to the current ARG
are inferred (allowing for recombination) using the
% chosen for each of the samples in the daily batch using the
Li and Stephens (LS) model (Section~\ref{sec:ls}). Then, since many samples
in the batch can share an attachment path,
we infer phylogenetic trees for each of the clusters separately using standard
methods (Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony-based heuristics to address issues
introduced by the inherent greediness of this strategy
(Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (with 1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of accurate joint estimation of
genetic inheritance with mutation and recombination events will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine in every generation,
in SARS-CoV-2 recombination is relatively rare, with recombination nodes accounting
for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of strains. We expect ARGs to be particularly tree-like
early in the pandemic, when co-infection was less likely and divergence between
lineages relatively low.

@@

% Paragraph 4, what do we say about the divergence times?
Figure~\ref{fig:recomb_mrcas} shows that there is a
large range in divergence times among recombinant parents, with a broad
spread of times from about 10 to 80 weeks prior to the recombination event
and a minor additional peak of recombinants that diverged
${\sim}95$ weeks, corresponding to recombinants whose
parental common ancestor traces to early 2020. This latter group should
contain, for example Delta-Omicron recombinants. More widely, we can
further classify the breakpoints by the VoC combinations of their
parent lineages. Considering only the Alpha, Delta, and Omicron VoC classes,
such a classification reveals that the majority of breakpoints have two Delta or
two Omicron parents, and that Omicron and Delta are the variants associated
with the most recombination (Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
This may reflect either sampling intensity, the
prevalence of cases (which increases the
chance of coinfection and recombination), or possible heterogeneity in
recombination probabilities among lineages.

@@

\section{Discussion}
% PARA 1. SARS-CoV-2 is still important, and the vast hordes of data provide an
% unprecedented opportunity to study the evolution of a virus. Even if the
% public health emergency is officially over, it doesn't mean it's not coming
% back and we should be prepared. Recombination is now a vital factor, and
% present a way in which it can be incorporated. Our results are accurate.
Although the COVID-19 pandemic is no longer considered a global emergency
by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide. This fuels
proliferation of many variants, with more than 600 Pango-designated lineages
circulating globally in the last three months (January to March, 2023;
https://gisaid.org/; access on March 27, 2023). High prevalence increases the risk of coinfection, providing opportunities for new phenotypically
distinct recombinants to emerge and spread. Phylogenetic approaches have
been central to pandemic responses thus far [FIXME: Examples and citations].
% 1. Form the basis of variant classifications (UShER paper),
% 2. Global context. In 2020, Nextstrain provided situation reports that were interactive, phylogenetically-driven visualizations of SARS-CoV-2 characteristics such as transmission https://nextstrain.org/ncov-sit-reps/
% 3. The WHO's recommendations on Omicron asked countries to "report initial cases/clusters associated with VOC infection to WHO". Phylogenetics being one way to identify those crucial early clusters. https://www.who.int/news/item/26-11-2021-classification-of-omicron-(b.1.1.529)-sars-cov-2-variant-of-concern
However, as recombinant lineages are now endemic and it is clear that
recombination is a major feature of ongoing SARS-CoV-2 evolution,
it is imperative that these methods are updated to incorporate the
effects of recombination, and future public health interventions are
not based on incomplete and potentially biased evolutionary models.
Here we have introduced the first method to infer an evolutionary
history that jointly estimates genealogies with both mutation and
recombination at pandemic scale, and illustrated how this single
structure accurately captures results derived by many different means.

@@

@@

The LS HMM machinery, and the interpretation of inferred recombinant paths and
breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
ways to extend and improve. For example, the current parameterisation of using
a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}), and likely
results in a flat likelihood space where many recombinant paths have equal
probability of being chosen.
Post-processing the match results to produce
more parsimonious breakpoints may also be a worthwhile avenue for development.
In particular, we may choose to insert breakpoints for the ARG that are chosen
from within the possible interval, rather than the current approach of taking
the rightmost value. Cases where we have more than two parents may either be
the ``stacking'' of multiple recombination events, or instances where
the HMM has chosen to switch to a third sequence rather than back to
the original parent (where this is equally parsimonious). Many putative
recombination events, however, will represent poor quality data, where
a recombinant copying path happens to be a more likely explanation
of a highly divergent sequence.
A thorough analysis of the behaviour of the LS model in the context
of a pandemic-scale ARG may lead to significant improvements in our
ability identify recombinants and to filter poor quality data.

@@

\section{Data Availability}
The source code for \texttt{sc2ts} and notebooks and code used to produce the
results described here available on GitHub:
\begin{itemize}
\item \url{https://github.com/jeromekelleher/sc2ts/}
\item \url{https://github.com/jeromekelleher/sc2ts-paper/}
\end{itemize}

@@
